D3064R0: Bringing the Laws of Physics, Computer-System Constraints, and Compilers to Bear on the OOTA Problem

### 2024-01-17 ootavacuous.tex
Fix signed/unsigned char example, add Gonzalo
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

\begin{quote}
	\co{void foo(char c)} \\
	\co{\{} \\
	\co{\ \ \ \ return c + 128;} \\
	\co{\}}
\end{quote}

The standard does not specify whether or not type \co{char} is signed,
leaving that choice to the implementations, which follow the lead
of the architects of a given CPU family.
Therefore, the code above is undefined behavior on systems choosing
signed \co{char} and having 8-bit bytes (so that there is no semantic
dependency from the \co{foo()} function's argument to its return value),
but perfectly well-defined on systems having unsigned \co{char}
(so that there is a semantic dependency).
So score one for the implementers, albeit by conscious choice on the
part of the standards committee.

@@

We are grateful to David Goldblatt, Jade Alglave, and Peter
Sewell for their careful review of an early draft of this paper and
to John Wickerson for asking Paul for a rant and taking the proffered
rant seriously.
We also owe David Goldblatt a debt of gratitude for his having asked an
insightful question at the right time, his insights on combinations of
OOTA and UB, and for his ``Deathstation 9000'' demonic CPU.
% Alan Stern located many unclear statements and gaps in reasoning.
Martin Uecker contributed valuable insights on backwards-propagating UB.
Richard Grisenthwaite patiently explained the architectural constraints
that prevent hardware OOTA.

>>>>

\begin{quote}
	\co{void foo(int x, char c)} \\
	\co{\{} \\
	\co{\ \ \ \ return x * (c >= 0);} \\
	\co{\}}
\end{quote}

The standard does not specify whether or not type \co{char} is signed,
leaving that choice to the implementations, which follow the lead
of the architects of a given CPU family.
Therefore, the code above has a a semantic dependency when run on systems
choosing signed \co{char}, but not when run on systems choosing unsigned
\co{char}.
So score one for the implementers, albeit by conscious choice on the
part of the standards committee.

@@

We are grateful to David Goldblatt, Jade Alglave, and Peter
Sewell for their careful review of an early draft of this paper and
to John Wickerson for asking Paul for a rant and taking the proffered
rant seriously.
We also owe David Goldblatt a debt of gratitude for his having asked an
insightful question at the right time, his insights on combinations of
OOTA and UB, and for his ``Deathstation 9000'' demonic CPU.
% Alan Stern located many unclear statements and gaps in reasoning.
Martin Uecker contributed valuable insights on backwards-propagating UB.
Gonzalo Brito Gadeschi shared an alternative way of handling non-volatile
atomic operations.
Richard Grisenthwaite patiently explained the architectural constraints
that prevent hardware OOTA.
